Closure-Free Functional Programming in a Two-Level Type Theory

### 2024-05-27 newpaper/paper.tex
revision wip

\newcommand{\vas}{\mathsf{as}}
\newcommand{\vbs}{\mathsf{bs}}
\newcommand{\vcs}{\mathsf{cs}}
\newcommand{\vxs}{\mathsf{xs}}
\newcommand{\vys}{\mathsf{ys}}
\newcommand{\vsp}{\mathsf{sp}}
\newcommand{\vma}{\mathsf{ma}}
\newcommand{\vm}{\mathsf{m}}
\newcommand{\vn}{\mathsf{n}}
\newcommand{\vk}{\mathsf{k}}
\newcommand{\vA}{\mathsf{A}}
\newcommand{\vB}{\mathsf{B}}
\newcommand{\vC}{\mathsf{C}}
\newcommand{\vS}{\mathsf{S}}
\newcommand{\vF}{\mathsf{F}}
\newcommand{\vR}{\mathsf{R}}
\newcommand{\vM}{\mathsf{M}}
\newcommand{\vmb}{\mathsf{mb}}
\newcommand{\mAs}{\mathsf{As}}
\newcommand{\va}{\mathsf{a}}
\newcommand{\vb}{\mathsf{b}}
\newcommand{\vc}{\mathsf{c}}
\newcommand{\vd}{\mathsf{d}}
\newcommand{\vx}{\mathsf{x}}
\newcommand{\vy}{\mathsf{y}}
\newcommand{\vz}{\mathsf{z}}
\newcommand{\vf}{\mathsf{f}}
\newcommand{\vfs}{\mathsf{fs}}
\newcommand{\vg}{\mathsf{g}}
\newcommand{\vh}{\mathsf{h}}
\newcommand{\vt}{\mathsf{t}}
\newcommand{\vs}{\mathsf{s}}
\newcommand{\vr}{\mathsf{r}}
\newcommand{\vu}{\mathsf{u}}
\newcommand{\vl}{\mathsf{l}}
\newcommand{\vns}{\mathsf{ns}}

@@

A pull stream is a meta-level specification of a state machine:
\begin{alignat*}{3}
  & \data \Step\,\vS\,\vA = \Stop\,|\,\Skip\,\vS\,|\,\Yield\,\vA\,\vS\\
  & \data \Pull\,(\vA : \MTy) : \MTy\,\where\\
  & \ind \Pull : (\vS : \MTy) \to \IsSOP\,\vS \RA \Gen\,\vS \to (\vS \to \Gen\,(\Step\,\vS\,\vA)) \to \Pull\,\vA
\end{alignat*}
In the $\Pull$ constructor, $\vS$ is the type of the internal state which is
required to be a sum-of-products of value types by the $\IsSumVS\,\vS$
constraint. Borrowing terminology from Coutts \cite{DBLP:phd/ethos/Coutts11}, we
call each product in the state a \emph{state shape}.

@@

How do we generate object code from streams? The state $\vS$ is given as a
finite sums of products, but the sums and the products are on the meta level, so
we cannot directly use $\vS$ in object code. Similarly as in the the treatment
of join points, we tabulate the $\vS \to \Gen\,\Step\,\vS\,\vA$ transition
function to a product of functions. However, these functions need to be
\emph{mutually recursive}, since it is possible to transition from any state to
any other state, and each such transition is represented as a function call.
This problem of generating well-typed mutual blocks was addressed by Yallop and
Kiselyov in \cite{DBLP:conf/pepm/YallopK19}. Unlike ibid., which used control
effects and mutable references in MetaOCaml, we present a solution that does not
use side effects in the metalanguage.

@@

We redefine the previous $\mit{Fun}_{\SOP\Lift}$ to return a computation type
instead:
\begin{alignat*}{3}
  &\mathrlap{\mit{Fun}_{\SOP\Lift} : \USOP \to \Ty \to \CTy}\\
  &\mit{Fun}_{\SOP\Lift}\,\Nil\,&& \vR = ()\\
  &\mit{Fun}_{\SOP\Lift}\,(\Cons\,\vA\,\vB)\,&& \vR = (\mit{foldr}\,(\to)\,\vR\,\vA,\,\mit{Fun}_{\SOP\Lift}\,\vB\,\vR)
\end{alignat*}
\begin{alignat*}{3}
  &\mtabulate &&: (\El_\SOP\,\vA \to \Lift \vR) \to \Lift(\mit{Fun}_{\SOP\Lift}\,\vA\,\vR)\\
  &\mindex    &&: \Lift(\mit{Fun}_{\SOP\Lift}\,\vA\,\vR) \to (\El_\SOP\,\vA \to \Lift \vR)
\end{alignat*}
Even for join points, this is just as efficient as the previous
$\mit{Fun}_{\SOP\Lift}$ version, since a definition of a product of functions
will get compiled to a sequence of function definitions. In addition, we can use
it to generate object code from streams. There are several choices for this, but
in CFTT the $\msf{foldr}$ function is as good as we can get.
\begin{alignat*}{3}
  &\hspace{-13em}\mathrlap{\foldr : \{\vA : \MTy\}\{\vB : \Ty\} \to (\vA \to \Lift \vB \to \Lift \vB) \to \Lift \vB \to \Pull\,\vA \to \Lift\,\vB}\\
  &\hspace{-13em}\mathrlap{\foldr\,\{\vA\}\,\{\vB\}\,(\Pull\,\vS\,\seed\,\step)\,\vf\,\vb = \ql}\\
  &\hspace{-13em}\mathrlap{\ind \letrec \vfs : \mit{Fun}_{\SOP\Lift}\,(\Rep\,\{\vS\})\,\vB := \spl\big(\mtabulate \dlr \lam \vs.\,\unGen\,(\step\,(\rep^{-1}\,\vs)) \dlr \lam \case} \\
  &\hspace{-13em}\ind \ind \ind \Stop            &&\to \vb \\
  &\hspace{-13em}\ind \ind \ind \Skip\,\vs       &&\to \mindex\,\qt{\vfs}\,(\rep\,\vs) \\
  &\hspace{-13em}\ind \ind \ind \Yield\,\va\,\vs &&\to \vf\,\va\,(\mindex\,\qt{\vfs}\,(\rep\,\vs))\big);\\
  &\hspace{-13em}\mathrlap{\ind \spl\big(\unGen\,\seed \dlr \lam s.\,\mindex\,\qt{\vfs}\,(\rep\,\vs)\big)\qr}
\end{alignat*}
This $\foldr$ is quite flexible becuse we can compute any object value from it,
including functions. For instance, we can define $\foldl$ from $\foldr$:
\begin{alignat*}{3}
  &\foldl : \{\vA : \MTy\}\{\vB : \VTy\} \to (\Lift \vB \to \vA \to \Lift \vB) \to \Lift \vB \to \Pull\,\vA \to \Lift\,\vB\\
  &\foldl\,\vf\,\vb\,\vas = \spl(\foldr\,(\lam \va\,\vg.\,\qt{\lam \vb.\,\spl \vg\,\spl(\vf\,\qt{\vb}\,\spl \va)})\,\qt{\lam \vb.\,\vb}\,\vas)\,\spl \vb
\end{alignat*}
Note that since we abstract $\vB$ in a runtime function, it must be a value
type. Here, each $\Stop$ and $\Yield$ in the transition function gets
interpreted as a $\lambda$-expression in the output. However, those $\lambda$-s
will be lifted out in the scope, yielding a proper mutually tail-recursive
definition with an accummulator for $\vB$. Contrast this to the GHC base
library, where $\foldl$ for lists is also defined from $\foldr$, to enable
push-based fusion, but where a substantial \emph{arity analysis} is used in
the compiler to (incompletely) eliminate the intermediate closures
\cite{DBLP:conf/sfp/Breitner14}.

@@

Can we have a list-like $\Monad$ instance for streams, with singleton streams
for $\return$ and $\concatMap$ for binding? This is not possible. Aiming at
\[ \concatMap : (\vA \to \Pull\,\vB) \to \Pull\,\vA \to \Pull\,\vB, \]
the $\vA \to \Pull\,\vB$ function can contain an infinite number of different
machine state types, which cannot be represented in a finite amount of object
code. Here by ``infinite'' we mean the notion that is internally available in
the meta type theory. For instance, we can define a $\Nat_\M \to \Pull\,\vB$
function which for each $\vn : \Nat_\M$ produces a concatenation of $\vn$
streams. Hence, we shall have the following function instead:
\[ \concatMap : \IsSOP\,\vA \RA (\vA \to \Pull\,\vB) \to \Pull\,\vA \to \Pull\,\vB \]
The idea is the following: if $\USOP$ is closed $\Sigma$-types, we can directly
define this $\concatMap$, by taking the appropriate dependent sum of the $\vA
\to \USOP$ family of machine states, which we extract from the $\vA \to
\Pull\,\vB$ function. Let us write $\Sigma\,\vA\,\vB : \MTy$ for dependent sums,
for $\vA : \MTy$ and $\vB : \vA \to \MTy$, and reuse $({,})$ for pairing.  We
also use the following field projection functions: $\mit{projS} : \Pull\,\vA \to
\MTy$, $\mit{projSeed} : (\vas : \Pull\,\vA) \to \mit{projS}\,\vas$, and
$\mit{projStep} : (\vas : \Pull\,\vA) \to \mit{projS}\,\vas \to
\Gen\,(\Step\,(\mit{projS}\,\vas)\,\vA)$.  For now, we assume that the following
instance exists:
\[ \instance (\IsSOP\,\vA,\,\{\va : \vA\} \to \IsSOP\,(\vB\,\va)) \RA \IsSOP\,(\Sigma\,\vA\,\vB) \]
Above, $\{\va : \vA\} \to \IsSOP\,(\vB\,\va)$ is a universally quantified
instance constraint; this can be also represented in Agda. The definition of
$\concatMap$ is as follows.
\begin{alignat*}{3}
  & \mathrlap{\concatMap : \IsSOP\,\vA \RA (\vA \to \Pull\,\vB) \to \Pull\,\vA \to \Pull\,\vB }\\
  & \mathrlap{\concatMap\,\{\vA\}\,\{\vB\}\,\vf\,(\Pull\,\vS\,\seed\,\step) = }\\
  & \mathrlap{\ind \Pull\,(\vS,\,\Maybe\,(\Sigma\,\vA\,(\mit{projS} \circ \vf)))\,(({,}) \fmap \seed\,\ap\,\return\,\Nothing) \dlr \lam \case }\\
  & \mathrlap{\ind \ind (\vs,\,\Nothing) \to \step\,\vs \bind \lam \case }\\
  & \ind \ind \ind \ind \Stop                 &&\to \return \Stop\\
  & \ind \ind \ind \ind \Skip\,\vs            &&\to \return \dlr \Skip\,(\vs,\, \Nothing)\\
  & \ind \ind \ind \ind \Yield\,\va\,\vs      &&\to \mdo \{\vs' \fro \mit{projSeed}\,(\vf\,\va); \return \dlr \Skip\,(\vs,\,\Just\,(\va,\,\vs'))\}\\
  & \mathrlap{\ind \ind (\vs,\,\Just\,(\va,\,\vs')) \to \mit{projStep}\,(\vf\,\va)\,\vs' \bind \lam \case }\\
  & \ind \ind \ind \ind \Stop                &&\to \return \dlr \Skip\,(\vs,\,\Nothing) \\
  & \ind \ind \ind \ind \Skip\,\vs'          &&\to \return \dlr \Skip\,(\vs,\,\Just\,(\va,\,\vs'))\\
  & \ind \ind \ind \ind \Yield\,\vb\,\vs'    &&\to \return \dlr \Yield\,\vb\,(\vs,\, \Just\,(\va,\,\vs'))
\end{alignat*}
Here, $\Nothing$ marks the states where we are in the ``outer'' loop, running
the $\Pull\,\vA$ stream until we get its next value. $\Just$ marks the states of
the ``inner'' loop, where we have a concrete $\va : \vA$ value and we run the
$(\vf\,\va)$ stream until it stops. In the inner loop, the machine state type
depends on the $\va : \vA$ value, hence the need for $\Sigma$. How do we get
$\IsSOP$ for $\Sigma$? The key observations are:
\begin{itemize}
 \item Metaprograms cannot inspect the structure of object terms.
 \item Object types do not depend on object terms.
\end{itemize}
Hence, we expect that during staging, every $\vf : \Lift\,\vA \to \VTy$ function
has to be constant. This is actually true in the staging semantics of
CFTT. There, all metafunctions are stable under object-level parallel
substitutions. Also, object types are untouched by substitution. Hence, a
straightforward unfolding of semantic definitions validates the constancy of
$\vf$.

@@

While $\Pull$ is not a monad, and hence also not a $\MonadGen$, we can still use
a monadic style of stream programming with good ergonomics. First, we need
singleton streams for ``returning'':
\begin{alignat*}{3}
  & \mit{single} : \vA \to \Pull\,\vA \\
  & \mit{single}\,\va = \Pull\,\Bool_\M\,\True \dlr \lam \vb.\,\return \dlr\mbf{if}\,\vb\,\mbf{then}\,\Yield\,\va\,\False\,\mbf{else}\,\Stop
\end{alignat*}
Now, we should use this operation with care, since it has two states and can
contribute to a code size blow-up. For example, $\concatMap$ over $\single$
introduces at least a doubling of the number of state shapes. Although
let-insertion and case splitting could be derived as $\concatMap$ over
$\single$, to avoid the size explosion we give more specialized definitions
instead. First, we define a helper function for binding a single $\IsSOP$ value.
\begin{alignat*}{3}
  & \hspace{-11em}\mathrlap{\mit{bind_{single}} : \IsSOP\,\vA' \RA \Lift \vA \to (\Lift \vA \to \Gen\,\vA') \to (\vA' \to \Pull\,\vB) \to \Pull\,\vB}\\
  & \hspace{-11em}\mathrlap{\mit{bind_{single}}\,\va\,\vf\,\vg =}\\
  & \hspace{-11em}\mathrlap{\ind \Pull\,(\Sigma\,\vA'\,(\mit{projS} \circ \vg))\,(\mdo \{\va' \fro \vf\,\va; \vs \fro \mit{projSeed}\,(\vg\,\va'); \return (\va',\,\vs)) \dlr \lam \case}\\
  & \hspace{-11em}\mathrlap{\ind \ind (\va',\,\vs) \to \mit{projStep}\,(\vg\,\va')\,\vs \bind \lam \case }\\
  & \hspace{-11em}\ind \ind \ind \Stop            &&\to \return \Stop\\
  & \hspace{-11em}\ind \ind \ind \Skip\,\vs       &&\to \return \dlr \Skip\,(\va',\,\vs) \\
  & \hspace{-11em}\ind \ind \ind \Yield\,\vb\,\vs &&\to \return \dlr \Yield\,\vb\,(\va',\,\vs)
\end{alignat*}
We recover let-insertion and case splitting as follows:
\begin{alignat*}{3}
  &\mit{gen_\Pull} : \Lift \vA \to (\Lift \vA \to \Pull\,\vB) \to \Pull\,\vB\\
  &\mit{gen_\Pull}\,\va = \mit{bind_{single}}\,\va\,\gen
\end{alignat*}
\begin{alignat*}{3}
  &\mit{case_\Pull} : (\mit{Split}\,\vA,\,\IsSOP\,(\mit{SplitTo}\,\{\vA\})) \RA \Lift \vA \to (\mit{SplitTo}\,\{\vA\} \to \Pull\,\vB) \to \Pull\,\vB\\
  &\mit{case_\Pull}\,\va = \mit{bind_{single}}\,\va\,\mit{split}
\end{alignat*}
Let us look at small example. We write $\mit{forEach}$ for $\concatMap$ with
flipped arguments.
\begin{alignat*}{3}
  &\hspace{-4em}\mathrlap{\mit{forEach}\,(\mit{take}\,\qt{100}\,(\mit{countFrom}\,\qt{0})) \dlr \lam \vx.}\\
  &\hspace{-4em}\mathrlap{\gen_\Pull\, \qt{\spl \vx * 2} \dlr \lam \vy.}\\
  &\hspace{-4em}\mathrlap{\mit{case_\Pull}\,\qt{\spl \vx < 50} \dlr \lam \case}\\
  &\hspace{-4em} \ind \True  &&\to \mit{take}\,\vy\,(\mit{countFrom}\,x)\\
  &\hspace{-4em} \ind \False &&\to \single\,\vy
\end{alignat*}
Here, in every $\forEach$ iteration, $\gen_\Pull\,\qt{\spl \vx * 2}$ evaluates
the given expression and saves the result to the machine state. Then,
$\mit{case_\Pull}$ branches on an object-level Boolean expression. If we use
$\foldr$ to generate object code from this definition, we get \emph{four}
mutually defined functions. We get two state shapes from $\single\,\vy$ and one
from $\mit{take}\,\vy\,(\mit{countFrom}\,x)$; we sum these to get three for the
$\mit{case_\Pull}$, then finally we get an extra shape from $\mit{forEach}$ which
introduces an additional $\Maybe$.

>>>>

\newcommand{\vas}{\mathsf{as}}
\newcommand{\vbs}{\mathsf{bs}}
\newcommand{\vcs}{\mathsf{cs}}
\newcommand{\vxs}{\mathsf{xs}}
\newcommand{\vys}{\mathsf{ys}}
\newcommand{\vsp}{\mathsf{sp}}
\newcommand{\vma}{\mathsf{ma}}
\newcommand{\vga}{\mathsf{ga}}
\newcommand{\vm}{\mathsf{m}}
\newcommand{\vn}{\mathsf{n}}
\newcommand{\vk}{\mathsf{k}}
\newcommand{\vA}{\mathsf{A}}
\newcommand{\vB}{\mathsf{B}}
\newcommand{\vC}{\mathsf{C}}
\newcommand{\vS}{\mathsf{S}}
\newcommand{\vF}{\mathsf{F}}
\newcommand{\vR}{\mathsf{R}}
\newcommand{\vM}{\mathsf{M}}
\newcommand{\vmb}{\mathsf{mb}}
\newcommand{\mAs}{\mathsf{As}}
\newcommand{\va}{\mathsf{a}}
\newcommand{\vb}{\mathsf{b}}
\newcommand{\vc}{\mathsf{c}}
\newcommand{\vd}{\mathsf{d}}
\newcommand{\vx}{\mathsf{x}}
\newcommand{\vy}{\mathsf{y}}
\newcommand{\vz}{\mathsf{z}}
\newcommand{\vf}{\mathsf{f}}
\newcommand{\vfs}{\mathsf{fs}}
\newcommand{\vg}{\mathsf{g}}
\newcommand{\vh}{\mathsf{h}}
\newcommand{\vt}{\mathsf{t}}
\newcommand{\vs}{\mathsf{s}}
\newcommand{\vr}{\mathsf{r}}
\newcommand{\vu}{\mathsf{u}}
\newcommand{\vl}{\mathsf{l}}
\newcommand{\vns}{\mathsf{ns}}

@@

A pull stream is a meta-level specification of a state machine:
\begin{alignat*}{3}
  & \data \Step\,\vS\,\vA = \Stop\,|\,\Skip\,\vS\,|\,\Yield\,\vA\,\vS\\
  & \data \Pull\,(\vA : \MTy) : \MTy\,\where\\
  & \ind \Pull : (\vS : \MTy) \to \IsSOP\,\vS \RA \Gen\,\vS \to (\vS \to \Gen\,(\Step\,\vS\,\vA)) \to \Pull\,\vA
\end{alignat*}
In the $\Pull$ constructor, $\vS$ is the type of the internal state which is
required to be a sum-of-products of value types by the $\IsSOP\,\vS$
constraint. Borrowing terminology from Coutts \cite{DBLP:phd/ethos/Coutts11}, we
call each product in the state a \emph{state shape}.

@@

How do we generate object code from streams? The state $\vS$ is given as a
finite sums of products, but the sums and the products are on the meta level, so
we cannot directly use $\vS$ in object code. Similarly as in the the treatment
of join points, we tabulate the $\vS \to \Gen\,\Step\,\vS\,\vA$ transition
function to a product of functions. However, these functions need to be
\emph{mutually recursive}, since it is possible to transition from any state to
any other state, and each such transition is represented as a function call.
This problem of generating well-typed mutual blocks was addressed by Yallop and
Kiselyov in \cite{DBLP:conf/pepm/YallopK19}. In contrast to this work, which
used control effects and mutable references in MetaOCaml, we present a solution
that does not use side effects in the metalanguage.

@@

We redefine the previous $\mit{Fun}_{\SOP\Lift}$ to return a computation type
instead:
\begin{alignat*}{3}
  &\mathrlap{\mit{Fun}_{\SOP\Lift} : \USOP \to \Ty \to \CTy}\\
  &\mit{Fun}_{\SOP\Lift}\,\Nil\,&& \vR = ()\\
  &\mit{Fun}_{\SOP\Lift}\,(\Cons\,\vA\,\vB)\,&& \vR = (\mit{foldr}\,(\to)\,\vR\,\vA,\,\mit{Fun}_{\SOP\Lift}\,\vB\,\vR)
\end{alignat*}
\begin{alignat*}{3}
  &\mtabulate &&: (\El_\SOP\,\vA \to \Lift \vR) \to \Lift(\mit{Fun}_{\SOP\Lift}\,\vA\,\vR)\\
  &\mindex    &&: \Lift(\mit{Fun}_{\SOP\Lift}\,\vA\,\vR) \to (\El_\SOP\,\vA \to \Lift \vR)
\end{alignat*}
Even for join points, this is just as efficient as the previous
$\mit{Fun}_{\SOP\Lift}$ version, since a definition of a product of functions
will get compiled to a sequence of function definitions. In addition, we can use
it to generate object code from streams. There are several choices for this, but
in CFTT the $\msf{foldr}$ function is as good as we can get.
\begin{alignat*}{3}
  &\hspace{-13em}\mathrlap{\foldr : \{\vA : \MTy\}\{\vB : \Ty\} \to (\vA \to \Lift \vB \to \Lift \vB) \to \Lift \vB \to \Pull\,\vA \to \Lift\,\vB}\\
  &\hspace{-13em}\mathrlap{\foldr\,\{\vA\}\,\{\vB\}\,\vf\,\vb\,(\Pull\,\vS\,\seed\,\step) = \ql}\\
  &\hspace{-13em}\mathrlap{\ind \letrec \vfs : \mit{Fun}_{\SOP\Lift}\,(\Rep\,\{\vS\})\,\vB := \spl\big(\mtabulate \dlr \lam \vs.\,\unGen\,(\step\,(\rep^{-1}\,\vs)) \dlr \lam \case} \\
  &\hspace{-13em}\ind \ind \ind \Stop            &&\to \vb \\
  &\hspace{-13em}\ind \ind \ind \Skip\,\vs       &&\to \mindex\,\qt{\vfs}\,(\rep\,\vs) \\
  &\hspace{-13em}\ind \ind \ind \Yield\,\va\,\vs &&\to \vf\,\va\,(\mindex\,\qt{\vfs}\,(\rep\,\vs))\big);\\
  &\hspace{-13em}\mathrlap{\ind \spl\big(\unGen\,\seed \dlr \lam s.\,\mindex\,\qt{\vfs}\,(\rep\,\vs)\big)\qr}
\end{alignat*}
This $\foldr$ is quite flexible because it can produce object terms with computation types.
For instance, we can define $\foldl$ from $\foldr$:
\begin{alignat*}{3}
  &\foldl : \{\vA : \MTy\}\{\vB : \VTy\} \to (\Lift \vB \to \vA \to \Lift \vB) \to \Lift \vB \to \Pull\,\vA \to \Lift\,\vB\\
  &\foldl\,\vf\,\vb\,\vas = \spl(\foldr\,(\lam \va\,\vg.\,\qt{\lam \vb.\,\spl \vg\,\spl(\vf\,\qt{\vb}\,\spl \va)})\,\qt{\lam \vb.\,\vb}\,\vas)\,\spl \vb
\end{alignat*}
Note that since we abstract $\vB$ in a runtime function, it must be a value
type. Here, each $\Stop$ and $\Yield$ in the transition function gets
interpreted as a $\lambda$-expression in the output. However, those $\lambda$-s
will be lifted out in the scope, yielding a proper mutually tail-recursive
definition with an accummulator for $\vB$. Contrast this to the GHC base
library, where $\foldl$ for lists is also defined from $\foldr$, to enable
push-based fusion, but where a substantial \emph{arity analysis} is used in
the compiler to (incompletely) eliminate the intermediate closures
\cite{DBLP:conf/sfp/Breitner14}.

@@

Can we have a list-like $\Monad$ instance for streams, with singleton streams
for $\return$ and $\concatMap$ for binding? This is not possible. Aiming at
\[ \concatMap : (\vA \to \Pull\,\vB) \to \Pull\,\vA \to \Pull\,\vB, \]
the $\vA \to \Pull\,\vB$ function can contain an infinite number of different
machine state types, which cannot be represented in a finite amount of object
code. Here by ``infinite'' we mean the notion that is internally available in
the meta type theory. For instance, we can define a $\Nat_\M \to \Pull\,\vB$
function which for each $\vn : \Nat_\M$ produces a concatenation of $\vn$
streams. Hence, we shall have the following function instead:
\[ \concatMap : \IsSOP\,\vA \RA (\vA \to \Pull\,\vB) \to \Pull\,\vA \to \Pull\,\vB \]
The idea is the following: if $\USOP$ is closed under $\Sigma$-types, we can directly
define this $\concatMap$, by taking the appropriate dependent sum of the $\vA
\to \USOP$ family of machine states, which we extract from the $\vA \to
\Pull\,\vB$ function. Let us write $\Sigma\,\vA\,\vB : \MTy$ for dependent sums,
for $\vA : \MTy$ and $\vB : \vA \to \MTy$, and reuse $({,})$ for pairing.  We
also use the following field projection functions: $\mit{projS} : \Pull\,\vA \to
\MTy$, $\mit{projSeed} : (\vas : \Pull\,\vA) \to \mit{projS}\,\vas$, and
$\mit{projStep} : (\vas : \Pull\,\vA) \to \mit{projS}\,\vas \to
\Gen\,(\Step\,(\mit{projS}\,\vas)\,\vA)$.  For now, we assume that the following
instance exists:
\[ \instance (\IsSOP\,\vA,\,\{\va : \vA\} \to \IsSOP\,(\vB\,\va)) \RA \IsSOP\,(\Sigma\,\vA\,\vB) \]
Above, $\{\va : \vA\} \to \IsSOP\,(\vB\,\va)$ is a universally quantified
instance constraint; this can be also represented in Agda. The definition of
$\concatMap$ is as follows.
\begin{alignat*}{3}
  & \mathrlap{\concatMap : \IsSOP\,\vA \RA (\vA \to \Pull\,\vB) \to \Pull\,\vA \to \Pull\,\vB }\\
  & \mathrlap{\concatMap\,\{\vA\}\,\{\vB\}\,\vf\,(\Pull\,\vS\,\seed\,\step) = }\\
  & \mathrlap{\ind \Pull\,(\vS,\,\Maybe\,(\Sigma\,\vA\,(\mit{projS} \circ \vf)))\,(({,}) \fmap \seed\,\ap\,\return\,\Nothing) \dlr \lam \case }\\
  & \mathrlap{\ind \ind (\vs,\,\Nothing) \to \step\,\vs \bind \lam \case }\\
  & \ind \ind \ind \ind \Stop                 &&\to \return \Stop\\
  & \ind \ind \ind \ind \Skip\,\vs            &&\to \return \dlr \Skip\,(\vs,\, \Nothing)\\
  & \ind \ind \ind \ind \Yield\,\va\,\vs      &&\to \mdo \{\vs' \fro \mit{projSeed}\,(\vf\,\va); \return \dlr \Skip\,(\vs,\,\Just\,(\va,\,\vs'))\}\\
  & \mathrlap{\ind \ind (\vs,\,\Just\,(\va,\,\vs')) \to \mit{projStep}\,(\vf\,\va)\,\vs' \bind \lam \case }\\
  & \ind \ind \ind \ind \Stop                &&\to \return \dlr \Skip\,(\vs,\,\Nothing) \\
  & \ind \ind \ind \ind \Skip\,\vs'          &&\to \return \dlr \Skip\,(\vs,\,\Just\,(\va,\,\vs'))\\
  & \ind \ind \ind \ind \Yield\,\vb\,\vs'    &&\to \return \dlr \Yield\,\vb\,(\vs,\, \Just\,(\va,\,\vs'))
\end{alignat*}
Here, $\Nothing$ marks the states where we are in the ``outer'' loop, running
the $\Pull\,\vA$ stream until we get its next value. $\Just$ marks the states of
the ``inner'' loop, where we have a concrete $\va : \vA$ value and we run the
$(\vf\,\va)$ stream until it stops. In the inner loop, the machine state type
depends on the $\va : \vA$ value, hence the need for $\Sigma$. How do we get
$\IsSOP$ for $\Sigma$? The key observations are:
\begin{itemize}
 \item Metaprograms cannot inspect the structure of object terms.
 \item Object types do not depend on object terms.
\end{itemize}
Hence, we expect that during staging, every $\vf : \Lift\,\vA \to \VTy$ function
has to be constant. This is actually true in the staging semantics of
CFTT. There, all metafunctions are stable under object-level parallel
substitutions. Also, object types are untouched by substitution. Hence, a
straightforward unfolding of semantic definitions validates the constancy of
$\vf$.

@@

While $\Pull$ is not a monad, and hence also not a $\MonadGen$, we can still use
a monadic style of stream programming with good ergonomics. First, we need
singleton streams for ``returning'':
\begin{alignat*}{3}
  & \mit{single} : \vA \to \Pull\,\vA \\
  & \mit{single}\,\va = \Pull\,\Bool_\M\,\True \dlr \lam \vb.\,\return \dlr\mbf{if}\,\vb\,\mbf{then}\,\Yield\,\va\,\False\,\mbf{else}\,\Stop
\end{alignat*}
Now, we should use this operation with care, since it has two states and can
contribute to a code size blow-up. For example, $\concatMap$ over $\single$
introduces at least a doubling of the number of state shapes. Although
let-insertion and case splitting could be derived as $\concatMap$ over
$\single$, to avoid the size explosion we give more specialized definitions
instead. First, we define an embedding of $\Gen$ operations in streams.
\begin{alignat*}{3}
  & \hspace{-11em}\mathrlap{\mit{bind_{\Gen}} : \IsSOP\,\vA \RA \Gen\,\vA \to (\vA \to \Pull\,\vB) \to \Pull\,\vB}\\
  & \hspace{-11em}\mathrlap{\mit{bind_{\Gen}}\,\vga\,\vf\,=}\\
  & \hspace{-11em}\mathrlap{\ind \Pull\,(\Sigma\,\vA\,(\mit{projS} \circ \vf))\,(\mdo \{\va \fro \vga; \vs \fro \mit{projSeed}\,(\vf\,\va); \return (\va,\,\vs)) \dlr \lam \case}\\
  & \hspace{-11em}\mathrlap{\ind \ind (\va,\,\vs) \to \mit{projStep}\,(\vf\,\va)\,\vs \bind \lam \case }\\
  & \hspace{-11em}\ind \ind \ind \Stop            &&\to \return \Stop\\
  & \hspace{-11em}\ind \ind \ind \Skip\,\vs       &&\to \return \dlr \Skip\,(\va,\,\vs) \\
  & \hspace{-11em}\ind \ind \ind \Yield\,\vb\,\vs &&\to \return \dlr \Yield\,\vb\,(\va,\,\vs)
\end{alignat*}
We recover let-insertion and case splitting as follows:
\begin{alignat*}{3}
  &\mit{gen_\Pull} : \Lift \vA \to (\Lift \vA \to \Pull\,\vB) \to \Pull\,\vB\\
  &\mit{gen_\Pull}\,\va = \mit{bind_{\Gen}}\,(\gen\,\va)
\end{alignat*}
\begin{alignat*}{3}
  &\mit{case_\Pull} : (\mit{Split}\,\vA,\,\IsSOP\,(\mit{SplitTo}\,\{\vA\})) \RA \Lift \vA \to (\mit{SplitTo}\,\{\vA\} \to \Pull\,\vB) \to \Pull\,\vB\\
  &\mit{case_\Pull}\,\va = \mit{bind_{\Gen}}\,(\mit{split}\,\va)
\end{alignat*}
Let us look at small example. We write $\mit{forEach}$ for $\concatMap$ with
flipped arguments.
\begin{alignat*}{3}
  &\hspace{-4em}\mathrlap{\mit{forEach}\,(\mit{take}\,\qt{100}\,(\mit{countFrom}\,\qt{0})) \dlr \lam \vx.}\\
  &\hspace{-4em}\mathrlap{\gen_\Pull\, \qt{\spl \vx * 2} \dlr \lam \vy.}\\
  &\hspace{-4em}\mathrlap{\mit{case_\Pull}\,\qt{\spl \vx < 50} \dlr \lam \case}\\
  &\hspace{-4em} \ind \True  &&\to \mit{take}\,\vy\,(\mit{countFrom}\,x)\\
  &\hspace{-4em} \ind \False &&\to \single\,\vy
\end{alignat*}
Here, in every $\forEach$ iteration, $\gen_\Pull\,\qt{\spl \vx * 2}$ evaluates
the given expression and saves the result to the machine state. Then,
$\mit{case_\Pull}$ branches on an object-level Boolean expression. If we use
$\foldr$ to generate object code from this definition, we get \emph{four}
mutually defined functions. We get two state shapes from $\single\,\vy$ and one
from $\mit{take}\,\vy\,(\mit{countFrom}\,x)$; we sum these to get three for the
$\mit{case_\Pull}$, then finally we get an extra shape from $\mit{forEach}$ which
introduces an additional $\Maybe$.
